---
title: "DMC Lesson 1 Models and Data"
---

*Note*: Before running these scripts, the current working directory must be set to the top-level folder containing the dmc
 and tutorial subfolders


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'C:/Code/dmc/')

```


# Lesson 1.1 Models

```{r include=FALSE}
rm(list=ls())
source ("dmc/dmc.R")
```

For this lesson we're using the LBA model. Only one model can be active at a time. Models are specified in the "models"
directory under the "dmc" directory. It also contains further sub-directories, the names of which are the first argument
to the load_model function. These subdirectories contain a file "dists.R" which defines the random and likelihood
functions that define the model in its core parameterization. The second argument to load_model specifies a particular
model parameterization. For the following LBA model the core parameterization is in terms of the threshold "b" parameter
but we will use a re-parameterization B=b-A (where A is start point variability) defined in "lba_B.R".

```{r include=FALSE}
load_model(dir_name="LBA",model_name="lba_B.R")
```

**NB1**: See ?LBA for rtdists package help on the LBA parameters.  
**NB2**: More advanced users can create their own reparameterizations using an
     existing model file as a template to alter the three functions it contains
     transform.dmc, random.dmc and likelihood.dmc (See the lessons in
     dmc_2_3_AddModel1.R and dmc_2_4_AddModel2.R, but it is recommended you
     work through to the end of lesson 4 first.


- SETTING UP A MODEL requires 6 steps, in three groups. 
- Steps 1-5 create a set of list and vector objects, and can be done in any order. 
- Step 6 uses these objects as arguments to the "model.dmc" function, which creates an array object ("model") with a set
  of attributes specifying a particular model and parameterization.

## Example 1

A single subject with parameters varying with response (R) and correctness (M, for match) in a minimal design with one
factor (S, a two level stimulus factor) that is used to score two possible responses named "r1" and "r2".

The first three steps specify the factors, possible responses, and how they are scored for accuracy (this mapping
creates a new factor "M", with levels "true" and "false").

The next two steps specify details about the model parameters and how they relate to factors (including M).

**(1)** Create a named list ( e.g., "factors" ) specifying factor names ( = list names ) and factor levels ( = character
vectors assigned to corresponding list elements ).

``` {r}
factors <- list(S=c("s1","s2"))  # Stimulus 1 and Stimulus 2
```


NB: Factor levels cannot contain a "." as this is used by DMC to
concantanate factor levels from multiple factors.


**(2)** Create a character vector ( e.g., "responses" ) giving the names of the
possible responses ( usually levels (data$R) in the data frame )


``` {r}
responses <- c("r1","r2")        # Response 1 and Response 2
```


**(3)** Create "match.map", a list of lists, where the first entry is a list named "M" that specifies which accumulator (or
boundary, for the DDM model) matches the correct response for each factor level. Accumulators/boundaries can be
specified by the corresponding response name or an integer 1..n for n response possibilities (see dmc_1_6 and dmc_1_7
for more advanced scoring when there are more than two responses).

``` {r}
match.map <- list(M=list(s1="r1",s2="r2"))
```

NB: Typically the levels all come from one factor, but can be a combination of levels from different factors. In this
case the levels must be specified in the same order as factor names (see 1 above and dmc_5_1 and dmc_5_2)

For this example we will setup an LBA model where the B parameter varies with the response (R), mean_v and sd_v vary
with match (M), and there are constants for st0 and for the sd_v of the mismatching accumulator.

**(4)** Create "p.map", a list which tells DMC how the factors specified above map onto the parameters of the model.
These "EXTERNAL PARAMETERS TYPES" are defined in a model parameterization file ("lba_B.R" in this example), and
"FACTORS" are specified in steps (1) & (3) (usually M for the latter).

``` {r}
p.map <- list(A="1",B="R",t0="1",mean_v="M",sd_v="M",st0="1")
```

NB1: The R and M factors MUST be given last when used with a parameter influenced by more than one factor

NB2: The "1" indicates that the same estimated value will be assumed for all cells in the design (i.e., an "intercept"
parameter).



**(5)** Create "const", a vector which specifies the parameters to be held constant (i.e., not varied in fitting), and their
values.

``` {r}
constants <- c(sd_v.false=1,st0=0)
```


NB: DMC names parameters by the parameter-type name and (possibly) concatenated factor levels to which they apply, e.g.,
"parameter_type_name.level_factor_A.level_factor_B ..." Again these levels must be in the same order as names in step
(1) Once a model is created you can view these names (assuming you called the model "model") using :
attr(model,"p.vector")


Now we create the model object, specifying the "norm" distribution function (one of the options for the LBA model in the
"rtdists" package "n1PDF" function, passed through to the "likelihood.dmc" function in the "lba_B.R" script.)


**(6)** Use the objects created in steps 1 to 5 as arguments to the function
"model.dmc" in order to create a Boolean array (with a set of attributes,
e.g., "model") that specifies the model to be fit.

``` {r}
model <- model.dmc(p.map,match.map=match.map,constants=constants,
                   factors=factors,responses=responses,type="norm")
model
```


NB1: The following things are checked and will cause a STOP:

1. Factors names "M", "R", and "1" (intecept factor) are reserved
2. Factor levels must be unique within AND BETWEEN factors
3. List "match.map" must have values in responses
4. Where match.map contains more than one list their names cannot be the same as factor names from (1)


This information is also available as attributes of the model object, e.g:

``` {r}
attr(model,"p.vector")
attr(model,"constants")
attr(model,"type")
attr(model,"posdrift")
```

NB: The "posdrift" boolean is a modifier specific to the norm type of the LBA model (see "rtdists")

## Model Matrix

There is one matrix for each accumulator, with row names showing the level.response naming and the columns the
parameter-type.level naming for parameters. The booleans show which parameters map to which design+response cell.

"model" is used by the function "p.df.dmc" (in dmc_model.R) to perform parameter assignment for each row (i) of the
model array, and transforms and creates parameters as specified in "transform.dmc" (in "lba_B.R")

To see this, first create a parameter vector to map. These are not very sensible values, chosen just so we can see
differences in the following printout

``` {r}
p.vector  <- c(A=3,B.r1=4,B.r2=5,
               mean_v.true=2,mean_v.false=-1,
               sd_v.true=0.5,
               t0=.2)
```

The "model" array above is computationally efficient but it is not easy to see how parameters map to design cells. The
function "print.cell.p" makes it easy to loop through cells and print the mapping.

``` {r}
print.cell.p(p.vector,model)
```


NB1: the rows are in "n1" order, that is, for ".r1" cells the first row is "r1", for ".r2" cells the first row is "r2".
This is because "n1PDF" (in the "rtdists" package) returns the likelihood for first unit or node in a set of
accumulators (the term n1 mean "node one") with parameters corresponding to the first entries in each parameter type's
parameter vector.

NB2: "transform.dmc" creates b = A+B, so the columns are the "internal" parameter types for the LBA (i.e., the name
expected by "n1PDF" given the type designator passed to likelihood.dmc). "transform.dmc" can also be used to change the
scale on which parameters are sampled. For example, you might sample sd_v on a log scale to enforce positivity, so
"transform.dmc" would have to calculate `par.df$sd_v <- exp(par.df$log_sd_v)` 

Most of the DMC lessons use very simple designs. If you want to know how to setup more complex designs see the advanced
lesson dmc_5_1_Factorial.R. If you want to see how to setup more complex scoring see dmc_5_2_Scoring.R.

# Lesson 1.2 Simulating one subject's data, and likelihoods

```
Note: Before running these scripts, the current working directory must be set 
to the top-level folder containing the dmc and tutorial subfolders 

RSTUDIO USERS: Plots can cause the error message: "Error in plot.new() : figure margins too large". This is not a DMC
bug, it just means your screen size is limited. You can fix this by maximizing the plot pane before plotting, or
reducing the number of panels and plotting multiple windows using par(mfrow=c(x,y)) with smaller x and y values (see
below for details).
```

It is essential to be able to simulate data from any model you want to use for several reasons:

1. So you can play with the model, changing its parameters and seeing the effects, just as you would run experiments and
analyse data to understand the effects of experimental manipulations.
2. So you can check your estimation procedure is working. If you can’t recover the parameters that you used to simulate
a very large data set data by fitting to that simulated data then something is wrong.
3. To check if your estimation procedure will work in the experimental design you intend to apply it to. Sometimes the
estimation procedure works in very large samples but fails in the design you are using in real a real experiment.
Simulating and fitting helps guide you to the design you need.


### Simulating Data.

The function `simulate.dmc` (in `dmc_model.R`) makes a data frame based on the parameter vector and the model, with _n_
observations for each row in model. _n_ can be a scalar or a vector of length equal to the number of rows (i.e., design
cells) in model.

``` {r}
data <- simulate.dmc(p.vector,model,n=1)
```

```
Indicative results (subject to randomness):
S  R       RT
1 s1 r1 3.308407
2 s2 r2 2.336722
```

The function `data.model.dmc` combines the data frame (which would normally come from an experiment rather than being
simulated) and the model so it can be passed as a single object to `likelihood.dmc` (in `lba_B.R` in `dmc/models`) in
order to calculate the likelihood of a parameter vector.

The likelihood summarizes all of the information in the data relevant to model fitting. As you will see in later lessons
having a likelihood is the key to being able to perform Bayesian estimation. By convention, we call a specific
combination of experimental data and a particular model a data-model instance (`data.model`)

``` {r}
data.model <- data.model.dmc(data,model)
```

This also adds "cell.index" and "cell.empty" attributes used to speed the computation of the likelihood.
`likelihood.dmc` computes a vector of likelihoods, one for each RT (in the) same order as in the data frame.
_Note that the output must always be greater than or equal to the function's parameter `min.like` argument (`1e-10` by default)._

``` {r}
likelihood.dmc(p.vector,data.model)
```

At this point it is worth taking a look at `lba_B.R`: `transform.dmc` allows a flexible relationship between the standard
model and the model parameterization (e.g., `b` vs. `B`) `likelihood.dmc` interfaces with the rtdists function

Let's simulate some more realistic data (with `n=1e4` observations per row):

``` {r}
p.vector <- c(A=1,B.r1=1,B.r2=1,mean_v.true=1,mean_v.false=0,sd_v.true=1,t0=.2)
data1 <- simulate.dmc(p.vector,model,1e4)
data.model1 <- data.model.dmc(data1,model)
```

Look at the accuracy and density for each stimulus level (correct responses are in black and error responses are in red)

``` {r}
plot.cell.density(data.model1[data.model1$S=="s1",],C="r1",xlim=c(0,4))
plot.cell.density(data.model1[data.model1$S=="s2",],C="r2",xlim=c(0,4))
```

Look at how summed log-likelihood varies around the true value using function `profile.dmc` in `dmc_plotting.R`. This
function takes a parameter name and min/max values to graph, and also prints the maximum likelihood estimate for the
parameter, which should tend to the true value for large samples

``` {r}
ylim=c(-35000,-32000)
par(mfrow=c(2,3))
profile.dmc("A",           .1,  2,p.vector,data.model1,ylim=ylim)
profile.dmc("B.r1",        .1,  2,p.vector,data.model1,ylim=ylim)
profile.dmc("mean_v.true", .1,  2,p.vector,data.model1,ylim=ylim)
profile.dmc("sd_v.true",   .1,  2,p.vector,data.model1,ylim=ylim)
profile.dmc("t0",          .01,.4,p.vector,data.model1,ylim=ylim)
```

NB1: Rstudio sometimes give an error if you try to plot too many panels in one plot. To avoid this you can sometimes
make the plotting area big before you run the command. You could also ask for fewer panels e.g., `par(mfrow=c(1,3))`
which will do one row of 3 panels over two pages, or you could shrink the margins around the plots, e.g., put
`mar=rep(2,4)` inside `par()` (this shrinks each of the four margins around a panel to a value of `2`, you can use
smaller or larger values).   
NB2: If you dont specify the ylim argument or set its value to NA `profile.dmc` will choose its
own scale, but this will likely differ across parameters so be careful when you compare graphs for different parameters.

### Simulating Data For Unbalanced Designs.

An unbalanced design can be simulated by making n a vector of the same length as the number of cells in the design. Here
we have 2 observations for `s1` and 4 for `s2`.

``` {r}
simulate.dmc(p.vector,model,n=c(2,4))
```

\pagebreak

# 1.3 Exploring the LBA model

Start with a simple design with a binary stimulus factor, S 
(e.g, left vs. right motion) and corresponding responses.  

```{r}
factors=list(S=c("left","right"))
responses=c("LEFT","RIGHT")
match.map=list(M=list(left="LEFT",right="RIGHT"))

```


NB1: For the LBA one of B, mean_v or sd_v must be fixed in at least one
   cell of the design. In this case we fix sd_v.false = 1. For simplicity we
   will also fix mean_v.false=0.
NB2: See ?LBA for rtdists help on parameters


```{r}
consts=c(st0=0,sd_v.false=1,mean_v.false=0)
```

Design where there are no effects except greater than chance accuracy due to
a `mean_v.true > mean_v.false` and Parameter vector names are: 

```{r}
p.map=list(A="1",B="1",mean_v="M",sd_v="M",t0="1",st0="1")
model <- model.dmc(type="norm",constants=consts,p.map=p.map,
                   match.map=match.map,factors=factors,responses=responses)
attr(model,"p.vector")

```

### 1) Fast and error prone performance

N.B. sd_v.true < sd_v.false as is often seen in practice

```{r}
p.vector  <- c(A=1,B=0,mean_v.true=1,sd_v.true=0.66,t0=.2)
data.model <- data.model.dmc(simulate.dmc(p.vector,model,n=1e4),model)

```


The score & plot:

```{r}
correct <- data.model$S==tolower(data.model$R)
round(mean(correct),2)                       # Accuracy
round(tapply(data.model$RT,list(correct),mean),2) # Errors/correct similar speed
plot.cell.density(data.model,C=correct,xlim=c(0,5)) # (red represents error)

```

### 2) Raise the threshold to improve accuracy, makes errors slow

Score & plot (using a convenience function which does the same as 
the previous code, useful for simple scoring situations such as this)

```{r}
p.vector  <- c(A=1,B=1,mean_v.true=1,sd_v.true=0.66,t0=.2)
data.model <- data.model.dmc(simulate.dmc(p.vector,model,n=1e4),model)
plot.score.dmc (data.model)
```





### 3) Lowering start-point noise has similar effects

```{r}
p.vector  <- c(A=.25,B=.75,mean_v.true=1,sd_v.true=0.66,t0=.2)
data.model <- data.model.dmc(simulate.dmc(p.vector,model,n=1e4),model)
plot.score.dmc (data.model)

```

NB: Further analyses are relative to case (2) above

### 4) Increasing matching rate increases accuracy and speed

```{r}
p.vector  <- c(A=1,B=1,mean_v.true=1.5,sd_v.true=0.66,t0=.2)
data.model <- data.model.dmc(simulate.dmc(p.vector,model,n=1e4),model)
plot.score.dmc (data.model)
```



### 5) Decreasing sd_v.true increases accuracy and creates fast errors

```{r}
p.vector  <- c(A=1,B=1,mean_v.true=1.5,sd_v.true=0.33,t0=.2)
data.model <- data.model.dmc(simulate.dmc(p.vector,model,n=1e4),model)
plot.score.dmc (data.model)
```


### 6) Increasing t0 slows things down again but has no effect on accuracy

```{r}
p.vector  <- c(A=1,B=1,mean_v.true=1.5,sd_v.true=0.66,t0=.4)
data.model <- data.model.dmc(simulate.dmc(p.vector,model,n=1e4),model)
plot.score.dmc (data.model)

```

*NB*: t0 has the same effect for all models.

Let's allow non-decision time to be variable across trials, so do not set `st0=0`.
Let's re-run `(2)` for comparison (but with a larger t0), and add a
printout of variability (as that is what's affected by st0). We use the inter-
quartile range to measure variability as it is robust against large outliers.


```{r}
p.vector  <- c(A=1,B=1,mean_v.true=1,sd_v.true=0.66,t0=.4)
data.model <- data.model.dmc(simulate.dmc(p.vector,model,n=1e4),model)
# Score & plot 
plot.score.dmc (data.model, IQR=TRUE)  # Show inter-quartile range

```

### 7) 


We set st0 to 0.2s. As t0 specifies the MINIMUM non-decision time we subtract st0/2 from t0 in order to maintain the
same mean non-decision time (what is usually referred to as ter). Also, any change in non-decision time parameters has
no effect on accuracy (because t0 distributes uniformly.)
    
```{r}
consts=c(sd_v.false=1,mean_v.false=0) # consts

p.map=list(A="1",B="1",mean_v="M",sd_v="M",t0="1",st0="1") # p.map

model <- model.dmc(type="norm",constants=consts,p.map=p.map,
                   match.map=match.map,factors=factors,responses=responses)

p.vector  <- c(A=1,B=1,mean_v.true=1,sd_v.true=0.66,t0=.1,st0=.2)
data.model <- data.model.dmc(simulate.dmc(p.vector,model,n=1e4),model)


# Score & plot
plot.score.dmc (data.model, IQR=TRUE)


```
    

### 8) Response bias can be introduced by allowing B to vary with accumulator (R)

```{r}

consts=c(st0=0,sd_v.false=1,mean_v.false=0)
p.map=list(A="1",B="R",mean_v="M",sd_v="M",t0="1",st0="1")
model <- model.dmc(type="norm",constants=consts,
                   p.map=p.map,match.map=match.map,
                   factors=factors,responses=responses)

```

For example, a bias to respond left
```{r}
p.vector  <- c(A=1,B.LEFT=.5,B.RIGHT=1.5,mean_v.true=1,sd_v.true=0.66,t0=.2)
data.model <- data.model.dmc(simulate.dmc(p.vector,model,n=1e4),model)
correct <- data.model$S==tolower(data.model$R)
# Right is now below chance
round(tapply(correct,data.model$S,mean),2)      
# Correct responses to left are faster and to right slower, but the opposite
# patterns shows in errors (and hence slow errors for left and fast for right)
round(tapply(data.model$RT,list(data.model$S,correct),mean),2)   
par(mfrow=c(1,2))
plot.cell.density(data.model[data.model$S=="left",],
                  C=correct[data.model$S=="left"],
                  xlim=c(0,5),ymax=1.0,main="left")
plot.cell.density(data.model[data.model$S=="right",],
                  C=correct[data.model$S=="right"],
                  xlim=c(0,5),ymax=1.0,main="right")

```


